---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, echo = TRUE, message = FALSE}
getwd()
raw <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r,echo = TRUE}
stepsDay <- aggregate(raw$steps, by = list(raw$date), FUN = sum, na.rm = TRUE)
names(stepsDay) <- c('Date', 'TotalSteps')
stepsDay
```
### Histogram of the total number of steps taken each day
```{r,echo = TRUE, message = FALSE}
require(ggplot2)
ts <- ggplot(stepsDay, aes(TotalSteps))
ts + geom_bar() + ylab("Number of days") + xlab("Total Number of Steps")
```

### mean and median of the total number of steps per day

```{r, echo = TRUE}
meanStepsDay <- mean(stepsDay$TotalSteps)
medianStepsDay <- median(stepsDay$TotalSteps)
meanMed <- data.frame(meanStepsDay, medianStepsDay)
meanMed
```

## What is the average daily activity pattern?
```{r, echo = TRUE}
avgInt <- aggregate(raw$steps, by = list(raw$interval), FUN = mean, na.rm = TRUE)
plot(x = avgInt$Group.1, y = avgInt$x, ylab = "Average Number of Steps", xlab = "5 Min Interval", type = "l" )
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo = TRUE}
maxStepInt <- subset(avgInt, x==max(avgInt$x))[1]
names(maxStepInt) <- c("Int. Start Minute")
maxStepInt
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset 
```{r, echo = TRUE}
dim(raw)[1] - sum(complete.cases(raw),na.rm = TRUE)
```

### Replace missing values in the dataset 

- Strategy: replace NA with mean steps for the interval
- calculate the mean for each interval 
- Duplicate the steps column for NA replacement (keep original steps column)
- Replace Na's in Duplicate steps column
- check 

```{r, echo = TRUE, message = FALSE}
require(plyr)
meanRaw <- ddply(raw,.(interval), summarize, mean=mean(steps,na.rm=TRUE))
raw$stepsNaR <- raw$steps
raw$stepsNaR[is.na(raw$stepsNaR)] <- meanRaw$mean[match(raw$interval,meanRaw$interval)][is.na(raw$stepsNaR)]
length(raw$stepsNaR[(is.na(raw$stepsNaR))])
```


### Create a new dataset with the missing data filled in

```{r, Echo = TRUE}
rawNaR <- raw[,-1]
names(rawNaR)[3] <- "steps"
str(rawNaR)
```

### Histogram of the total number of steps taken each day (NA's replaced)

```{r, Echo = TRUE, message = FALSE}
stepsDayNaR <- aggregate(rawNaR$steps, by = list(rawNaR$date), FUN = sum, na.rm = TRUE)
names(stepsDayNaR) <- c('Date', 'TotalSteps')
require(ggplot2)
ts2 <- ggplot(stepsDayNaR, aes(TotalSteps))
ts2 + geom_bar() + ylab("Number of days") + xlab("Total Number of Steps")
```

### Mean and median total number of steps taken per day after NA replacment  

```{r, Echo = TRUE, message = FALSE}
meanStepsDayNaR <- mean(stepsDayNaR$TotalSteps)
medianStepsDayNaR <-median(stepsDayNaR$TotalSteps)
mComp <- data.frame(mean = c(meanStepsDay,meanStepsDayNaR), 
                    median = c(medianStepsDay,medianStepsDayNaR),
                    row.names = c("With NAs", "NAs Replaced"))
mComp
```
### Do these values differ from the estimates from the first part of the assignment? 

The mean increased from 9,354 to 10,766. Interestingly the mean and median are the same once Na's are replaced.

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r, echo = TRUE}
totalStepsAdded <- sum(stepsDayNaR$TotalSteps) - sum(stepsDay$TotalSteps)
percIncrease <- (sum(stepsDayNaR$TotalSteps) - sum(stepsDay$TotalSteps))/sum(stepsDay$TotalSteps)
NaRImpact <- data.frame(totalStepsAdded, percIncrease)
NaRImpact
 
```
Replacing the Na's' added 86k steps, or 15%

## Are there differences in activity patterns between weekdays and weekends?

### Create a 2 level factor variable: “weekday” and “weekend” 

```{r, echo = TRUE}
rawNaR$Weekday <- ifelse(weekdays(as.Date(rawNaR$date)) %in% c("Saturday", "Sunday"), "Weekend", "Week")
```

### Panel plot: Weekend vs weekday average steps per interval 

```{r, echo = TRUE, message = FALSE}
require(ggplot2)
meanRawNaRWk <- ddply(rawNaR,.(interval, Weekday), summarize, 
                           mean=mean(steps,na.rm=TRUE))
g <- ggplot(meanRawNaRWk, aes(interval, mean))
g + geom_line() + facet_wrap(~Weekday, ncol=1) 
